Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-05-27 main.tex
Fixed a caption
Most of Allen's changes applied

\def\cvprPaperID{257} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

@@

\section{Introduction} 
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
$\ell_1$-min refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

For the completeness of the paper, in this section, we briefly describe the ALM
algorithm for $\ell_1$-min \eqref{eq:l1min} \cite{YangA2010-ICIP} and analyze
its complexity. Parallelization of ALM for face alignment and recognition will
be discussed in Section \ref{sec:alignment} and Section \ref{sec:recognition},
respectively.

@@

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
sizes and speeds of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
Sections \ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that while typical engineering
workstations can only support a single high-end GPU due to their high current
draw, purpose-built servers can support multiple high-end GPU's.  Unless
otherwise specified, all implementations utilize single precision floating
point datatypes.

\subsection{CPU Hardware Concurrency} 
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speeds.

@@

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\label{fig:caches}
\caption{Scale drawings for the visual comparison of the algorithm
working set to available caches}
\end{figure}

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KB L1 data cache and a private 256\,KB L2 cache.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8\,MB of L3 cache that is shared by the four cores.
Overall, the algorithm has approximately 16\,MB of L3 cache
available for a dual-processor configuration.

For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision 
floating point values simultaneously \footnote{By the publication date, the vector
units of typical high-end CPUS will double.}.

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em sse-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities\footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}.

@@

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache, and all SMP's share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:gpu_caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}.  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em warp-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

To mitigate the problem, if the system can obtain to a good initial guess of the transformation $\tau$ (\eg, provided by a good face detector),
the optimal solution for $\tau$ can be sought iteratively by a sequence of linearized approximations $j=1, 2, \cdots$ as follows:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.

The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:alignment_alm}. In terms of computational complexity, one significant
term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
computational cost is bounded by $O(n_i^2m + n_i^3)$. For per-class face
alignment, since both $A_i$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
for all $i=1, \cdots, K$ classes), we find that the computation time is still
dominated by the iterative computation of matrix-vector multiplications in the
inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:alignment_alm}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow (B_j^\dagger)^T \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau^*\leftarrow \tau_j, \e^*\leftarrow \ee_j$
\end{algorithm}

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  Therefore, if eight instances of
\eqref{eq:alignment-linearization} were be executed in parallel on the eight
cores, there would be an insufficient amount of L1 or L2 cache to hold $A$ and
$A^\dagger$ during the iterative process to execute Algorithm
\ref{alg:alignment_alm}. Therefore, the bottleneck in this strategy becomes the
bandwidth between the L2 cache and L3 cache. The implementation of some
previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this
mapping strategy.

In this paper, we show that for CPU parallelism of face alignment, a higher
performance implementation can be achieved by {\em not} taking advantage of
problem-level parallelism at all, and instead mapping the pixel-level parallelism
onto both the core-level and the sse-level concurrency provided by the CPU.  In
other words, the proposed CPU implementation performs alignment of a query
image against the training images of a single user at a time using all cores on
both processors.

@@

Solving one alignment problem per SMP necessitated the implementation of a
custom CUDA matrix inversion routine.  This was done using Gaussian elimination
with partial pivoting.  To achieve precision comparable to the single precision
LAPACK routines in intel's MKL with this simplistic algorithm, we use double
precision.  Since the matrix to be inverted is only $n \times n$, the cost of
the inversion is dwarfed by the cost of other steps.  Similarly, while a
commercial library might perform reductions on long vectors (i.e. sum or
$\ell_1$ norm of an image) using a tree structure to improve precision without
the cost of double precision math, we simply use a double precision temporary
variable for accumulation.

In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
After the face alignment stage is complete, a recognition stage consisting of a
single $\ell_1$ minimization is performed to represent the aligned testing
image as a sparse combination of aligned training images from the 20 users with lowest alignment
residual, as shown in \eqref{eq:l1min_denoise}. 
The coefficients $\x$ are then used to compute error residuals that are
used for classification.  This recognition algorithm has been extensively studied in 
\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.

The $ell_1$ minimization is performed via the ALM algorithm shown in 
Algorithm \ref{alg:alm_rec}.  In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit exploit pixel-level parallelism
on CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

\subsection{Recognition Stage Implementation} Compared to problem-level
parallelism, pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively easy to exploit using existing BLAS libraries.  For this reason,
both the CPU and GPU implementation of the recognition stage is very similar to
the CPU implementation of the $\ell_1$ minimization advocated in Section
\ref{sec:alignment_implementation_cpu}.  On the CPU, most of the operations map
well onto MKL BLAS calls, and the operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  All temporary variables are stored in GPU DRAM
during solver execution, with the sole exception of the pseudoinverse of A,
which is computed on the CPU via Intel's LAPACK implementation, and then
uploaded to GPU DRAM.

\section{Experiments} \label{sec:experiment} In this section, we benchmark the
performance of our parallel implementations of $\ell_1$ minimization on CPU and
GPU platforms.  In order to show how our $\ell_1$ minimization algorithms scale
with problem size, in Section \ref{sec:simulation} we will begin with
benchmarks for the general $\ell_1$-min problem on synthetic data.  We will
then progress in Sections \ref{sec:alignment_benchmark} and
\ref{sec:recognition_benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.

@@

The first experiment compares the performance of our two proposed
implementations of the general $\ell_1$-min (Algorithm \eqref{alg:alm}) given a
generic random matrix $A$ and a random sparse vector $\xx_0$.  The measurement
$d \times n$ matrix $A$ was generated as a random Gaussian matrix, with each
entry randomly generated from the standard normal and normalized with unit
column norm.  The ratio of $d/n$ =$1/2$ with $n$ varying from 1000 to 8000.
The ground truth signals, $\xx_0$ had a sparsity rate of 10\% of d with
elements generated using the standard normal and also normalized with unit
norm.  Because the ground truth signal $\xx_0$ was known, the algorithm
terminated when $\|\bx-\bx_0\| < \tau$ with $\tau=10^{-3}$.  The measurement
vector was generated with the equation $\bb_0 = A \xx_0$.

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the CPU implementation is faster at solving a
single small problem.  
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Speed vs Dimensions of $A$.}
\label{fig:random_data}
\end{figure}

\subsection{Face Alignment Stage Benchmark} 
\label{sec:alignment_benchmark}
This section presents benchmarks of our CPU and GPU implementations of the
alignment stage of the face recognition pipeline described in Algorithm
\ref{alg:iterative_alignment}.

First, in order to test our theory that the GPU performs best solving many
$\ell_1$ problems-per concurrently, we benchmarked three implementations of the
alignment $ell_1$ minimization with $A$ of size $5120 \times 32$, and a fixed
$50$ inner loop iterations for each of $50$ outer loop iterations. The runtime
on a GTX480 GPU was averaged over a large number of trials, which were run
sequentially or concurrently depending on the implementation.
%The results are shown in Table \ref{tbl:ubench}.
The results are shown below:
%\begin{table}[h] 
\begin{tabular}{|l|c|}
\hline
Sequential solver using CUBLAS & 302\,ms \\
\hline
One problem per SMP using CUDA streams 
\footnote{The streams implementation was limited significantly by a 
cap on the number of concurrent streams in the current version of CUDA}
& 70\,ms \\
\hline
One problem per SMP using single kernel & {\bf 36\,ms} \\
\hline
%\label{tbl:ubench}
\end{tabular}
%\end{table}
Our proposed parallelization of the $\ell_1$ minimization used in the alignment
stage is {\em eight times} faster than an implementation solving a single
problem at a time using the stock BLAS libraries.

>>>>

\def\cvprPaperID{184} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

@@

\section{Introduction} 
$\ell_1$ minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
$\ell_1$-min refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

For the completeness of the paper, in this section, we briefly describe the ALM
algorithm for $\ell_1$-min \eqref{eq:l1min} \cite{YangA2010-ICIP}
 %and analyze its complexity
. 
Parallelization of ALM for face alignment and recognition will
be discussed in Section \ref{sec:alignment} and Section \ref{sec:recognition},
respectively.

@@

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have big performance implications.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance implications to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. 
%GPUs are commonly shipped with a
%single GPU chip per PCI-express card.  Note that while typical engineering
%workstations can only support a single high-end GPU due to their high current
%draw, purpose-built servers can support multiple high-end GPU's.  
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.

\subsection{CPU Hardware Concurrency} 
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
cache, and relatively high clock speeds.

@@

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\label{fig:caches}
\caption{A visual comparison of the algorithm working set to the CPU and GPU
caches.  Note: Although the aspect ratio could not be preserved, all arrays and
caches are drawn so that the area is proportional to the size of the data. }
\end{figure}

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KiB L1 data cache and a private 256\,KiB L2 cache.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8\,MiB of L3 cache that is shared by the four cores.
Overall, the algorithm has approximately 16\,MiB of L3 cache
available for a dual-processor configuration.

For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision 
floating point values simultaneously.
There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em sse-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities\footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}.

@@

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KiB of L1 cache, and all SMP's share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:gpu_caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}.  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

To mitigate the problem, if the system can obtain a good initial guess of the
transformation $\tau$ (\eg, provided by a good face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step as follows:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:alignment_alm}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:alignment_alm}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow (B_j^\dagger)^T \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  Therefore, if eight instances of
\eqref{eq:alignment-linearization} were be executed in parallel on the eight
cores, there would be an insufficient amount of L1 or L2 cache to hold $B_j$ and
$B_j^\dagger$ during the iterative process to execute Algorithm
\ref{alg:alignment_alm}. Therefore, the bottleneck in this strategy becomes the
bandwidth between the L2 cache and L3 cache. The implementation of some
previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this
mapping strategy.

In this paper, we show that for CPU parallelism of face alignment, a higher
performance implementation can be achieved by {\em not} taking advantage of
problem-level parallelism at all, and instead mapping the pixel-level parallelism
onto both the core-level and the SSE-level concurrency provided by the CPU.  In
other words, the proposed CPU implementation performs alignment of a query
image against the training images of a single user at a time using all cores on
both processors.

@@

Solving one alignment problem per SMP necessitated the implementation of a
custom CUDA routine for computing $B_j^\dagger = (B_j^TB_j)^-1 B^T = G^-1 B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Additionally, reductions on long vectors (i.e. sum or $\ell_1$ norm
of an image) are performed using a double precision temporary variable for accumulation.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
After the face alignment stage is complete, all but the 20 users with lowest
$||\e_i||_1$ are thrown out, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

\subsection{Recognition Stage Implementation} Compared to problem-level
parallelism, pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively easy to exploit using existing BLAS libraries.  For this reason,
both the CPU and GPU implementation of the recognition stage is very similar to
the CPU implementation of the $\ell_1$-min advocated in Section
\ref{sec:alignment_implementation_cpu}.  On the CPU, most of the operations map
well onto MKL BLAS calls, and the operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  All temporary variables are stored in GPU DRAM
during solver execution, with the sole exception of the pseudo-inverse of $B_j$,
which is computed on the CPU via Intel's LAPACK implementation, and then
uploaded to GPU DRAM.

\section{Experiments} \label{sec:experiment} In this section, we benchmark the
performance of our parallel implementations of $\ell_1$-min on CPU and GPU
platforms.  In order to show how our $\ell_1$-min algorithms scale with problem
size, in Section \ref{sec:simulation} we will begin with benchmarks for the
general $\ell_1$-min problem on synthetic data.  We will then progress in
Section \ref{sec:benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.

@@

The first experiment compares the performance of our two proposed
implementations of the general $\ell_1$-min (Algorithm \eqref{alg:alm}) given a
generic random matrix $A$ and a random sparse vector $\xx_0$.  The measurement
$d \times n$ matrix $A$ is generated as a random Gaussian matrix, with each
entry randomly generated from the standard normal and normalized with unit
column norm.  The ratio of $d/n$ is fixed at $1/2$ with $n$ varying from 1000 to 8000.
The ground truth signals, $\xx_0$ have a sparsity rate of 10\% of d with
elements sampled from a normal distribution and normalized to unit $\ell_2$ vector
norm.  Because the ground truth signal $\xx_0$ is known, the algorithm
terminats when $\|\bx-\bx_0\| < \tau$ with $\tau=10^{-3}$.  The measurement
vector is generated with the equation $\bb_0 = A \xx_0$.

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the CPU implementation is faster at solving a
single small problem.  
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Runtime vs. Dimensions of $A$.}
\label{fig:random_data}
\end{figure}

\subsection{Face Alignment Stage Benchmark} 
\label{sec:benchmark}
This section presents benchmarks of our CPU and GPU implementations of the
alignment stage of the face recognition pipeline, as shown in Algorithm
\ref{alg:iterative_alignment}, as well as the recognition stage of the
face recognition pipeline, as seen in Algorithm \ref{alg:alm_rec}.

First, in order to test our theory that the GPU performs best solving many
$\ell_1$ problems-per concurrently, we benchmark three implementations of the
alignment $ell_1$-min with $A$ of size $5120 \times 32$, and a fixed
$50$ inner loop iterations for each of $50$ outer loop iterations. The runtime
on a GTX480 GPU is averaged over a large number of trials, which are run
sequentially or concurrently depending on the implementation.
The results are shown in Table \ref{tbl:ubench}.
%The results are shown below:
\begin{table}[h] 
\begin{tabular}{|l|c|}
\hline
Sequential solver using CUBLAS & 302\,ms \\
\hline
One problem per SMP using CUDA streams 
\footnote{The streams implementation is limited significantly by a 
cap on the number of concurrent streams in the current version of CUDA}
& 70\,ms \\
\hline
One problem per SMP using single kernel & {\bf 36\,ms} \\
%\hline
\label{tbl:ubench}
\end{tabular}
\end{table}
Our proposed parallelization of the $\ell_1$-min used in the alignment
stage is {\em eight times} faster than an implementation solving a single
problem at a time using the stock BLAS libraries.
